\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-03-04 review/response-B.Rmd

---
title: "Response"
author: "H. Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas Langrené, Patricia Menéndez"
date: '2023-03-04'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Response to B-review_1_1

## The Paper

The package and the proposed data structures could certainly constitute a valuable base for the spatio- temporal statistics tooling. Unfortunately the paper does not clearly describe the data structure, nor the associated functionality. I had to read the package vignettes, run the attached code and even look into the package’s code in order to fully understand the data structure.

**Re: **

Given the small size of the package, the paper feels a bit on the heavy side. Most of the code in the paper is not self-contained which makes it rather difficult to follow. Examples at the end of the paper don’t elucidate the inner workings of the package but rather advertise the functionality of other, mostly visualization, packages.

**Re: **

More fundamentally, the design of the package has an immediately apparent drawback. The cubble class is used to represent two incompatible data types - long and nested. Internally, the two data types are distinguished by the internal attribute “form”. In other words, the authors use an attribute of an object to mimic the functionality of a sub-class.

**Re: **

Relatedly, the rationale for the parallel naming convention long/nested and temporal/spatial is not entirely clear. It seems to me that the semantically unambiguous temporal/spatial could be used throughout without ambiguity, thus resolving the terminological redundancy.

**Re: **

My recommendation would be a combination of the following:

  * Rework the definitional parts (sections 2 and 3) by following more closely the “design” vignette. **Re: **
  * Describe at a glance the core functions which operate on the class and their motivation. **Re: **
  * Rework section 4 by making it more concise and provide links to sections in supplementary material or dedicated vignettes. **Re: **
  * Rework examples section 5 into “Applications”. Briefly describe the applications and provide and refer to dedicated and self-contained vignettes. **Re: **

## The Package

The package is generally badly documented. The meaning of the arguments is almost never clear from the documentation alone. For example the documentation of as_cubble states:

* What objects are supported as input data?
* What is the assumption of the “nestedness” of the input data. 
* What is the accepted data type of key, index and coords?

**Re: **

## Further suggestion

Consider not throwing when face_spacial is applied to the spatial object. This would allow for generic code where the “form” of the input the data is not known or does not matter.

**Re: **

It would be useful to print a few rows of the spatial attribute of the “long” object.

**Re: **

Section 3.6 is missing `tidyr` package which is especially relevant here given its nesting and unnesting operations https://tidyr.tidyverse.org/articles/nest.html

**Re: **

More comments inlined in the pdf.

  - on p4 "The following sections explain their roles, why the new cubble class is needed
and how the package relates to existing packages for spatial and temporal data analysis." The paper is a bit on the verbose side. Such comments could be safely removed. **Re:** sentence removed.

  - on p5 section 3.1, "The arguments key and index follow the wording in the tsibble package to describe the temporal order and multiple series while coords specifies the spatial location of each site.", Unclear what `key` and `index` mean. Isn't key a spatial identifier and index a temporal one? **Re: **

  - on p5 section 3.1, the word "wording" is highlighted in the sentence "The arguments key and index follow the wording in the tsibble package to ...", conventions. **Re:** *wording* changed to *conventions*

  - on p6 section 3.2, 'The cubble class also provides a long form, which expands the ts column and temporarily “hides” the spatial variables.', confusing: `hides` means storing it as an attribute. **Re: **

  - on p6 section 3.2, I think it would be a good idea to clearly describe the two data structures in an earlier section. By this point it's still not clear what those are and what the meaning of key and index is. **Re: **

  - on p6, section 3.2, yet another reason to call this data structure "cubble_temporal" instead of the semantically unobvious "long". **Re: **

  - on p8 section 3.5, Isn't the root reason the memory efficiency? That is, the spatial component cannot be represented as part of the nested tibble without repetition. **Re: **

  - on p8 section 3.6, The missing elephant is tidyr which has nesting and unnesting operations  https://tidyr.tidyverse.org/articles/nest.html. **Re: **

# Response to B-review_2_1

## The Manuscript

The manuscript is in general clearly structured. However, an important reference to the R-package sftime is missing that also deals with the representation of spatiotemporal data. It might be beneficial to provide illustrative examples in the manuscript that clearly compare between a cubble representation and representations of existing packages. Which data sets/structures cannot (or with a greater effort) be represented with existing R-packages such as e.g. stars and sftime?

**Re: **

For spatial data, the coordinate reference system is essential metadata information. In one example of the manuscript, the input data does have information on the CRS but it is not discussed in the manuscript how that is handled within cubble. How would coordinate transformations be handled with cubble?

**Re: **

The use-case of temporal matching based on features of the time series appears an interesting, but also a very specific one. How could this be generalized to a more generic approach?

**Re: **

Additionally, several typos and language issues arise and limit the readability of the manuscript. Some examples are:

- doubled/missing words/none correct sentences: 
  - “*components* spatio-temporal *components*” 
  - “... fits works ...” 
  - “... be activate rows ...” 
  - “... highlighted *in* the ...” 
  - “. . . using 2020 measurements using match_sites() function.” 
  - “An example of this using is included in the Appendix” 
  - “. . . the data in -a- multiple -of- ways on-the-fly” 
  - **Re**: the mentioned examples have been changed to: "... on different spatio-temporal components", "The workflow ... works well with ...", "... will activate rows having ...", "... highlighted in the map", "... using the match_sites() function.", "An example of this is included in the Appendix", and "... the data in multiple ways on-the-fly".

- Surprising references: in Section 4.4 it says “. . . Glyph maps (Section 3.4)”, but Section 3.4 says “. . . glyph maps will be explained in Section 4.4”. **Re:** the reference to Section 3.4 in Section 4.4 has been replaced with the literature (Wickham et al. 2012).

- Should “. . . it’s temperatures are more *consistent*” rather be “. . . it’s temperatures are more *constant*”? **Re: ** changed to "more constant"

- Typos:
  - “. . . spatial and *tmeporal* information are available.” 
  - The polar vortex, signalled by the high *speicfic* humidity, splits into two on 2002-09-26 and *further-s- split_s_* into four on 2002-10-04. 
  - the data in *-a- multiple -of-* ways on-the-fly. 
  - **Re: ** typo corrected to "temporal", "specific", "further splits", and "multiple ways"

## The package

I have been a bit puzzled by the “print” of a nested cubble in its temporal face. The row # temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl] appears at first sight (also in comparison with the print of the spatial face), as if the temporal domain is given by all those variables. To me, a notion such as

`# temporal: date [date]`

`# variables: prcp [dbl], tmax [dbl], tmin [dbl]`

would have been more intuitive. Possibly also including the temporal range (as for the spatial domain its bbox).

**Re: **

In the manuscript and in the reproducible-script.R, a data set `historical_tmax` is introduced. In the package, the corresponding data set seems to be `tmax_hist`.

**Re: **

The manuscript cannot be reproduced, as the reproducible-script.R (part of the submission to JSS) only uses a subset, but does not provide a separate result file that would allow to compare the reproduced outputs with the desired output for the subset.

**Re: **

### 2023-03-25 3-the-cubble-pkg.Rmd
some outlines on comparing between cubble/ sftime/stars

>>>>

## Comparison with existing packages

The development on spatio-temporal data structure in R has witnessed great changes from the initial \pkg{spacetime} [@spacetime] package. New packages, such as \pkg{stars} [@stars] and \pkg{sftime} [@sftime], have [...].

Below compare the additional benefit of cubble on top of the two packages.

The `stars` package creates a `stars` object from an array and a dimension object. In many cases, i.e. satellite images, this is convenient, however, there are also many spatio-temporal data are formatted in 2D data frames.

This requires converting the variables into an array format.

Consider the following spatial and temporal data.

The temporal data contains two variables are recorded at two stations (`id`) from 2023 March 21st to 2023 March 25th. Notice is that id2 only records from 2023 March 22nd and variable 2 has an `NA` value on  2023 March 23rd at id2.

```{r echo = FALSE}
library(stars)
spatial <- tibble(
  id = c(1,2),
  long = c(130, 140),
  lat = c(-38, -44),
  name = c("station 1", "station 2"))

spatial

temporal <- tibble(
  id = c(rep(1, each = 5), rep(2, each = 4)),
  time = c(seq(as.Date("2023-03-21"), as.Date("2023-03-25"), 1),
           seq(as.Date("2023-03-22"), as.Date("2023-03-25"), 1)),
  var1 = c(3, 6, 8, 2, 5, 7, 9, 14, 5),
  var2 = c(1, 3, 4, 2, 6, 7, NA, 5, 8)
)
temporal
```

The following code can be used to create the dimension object:

```{r echo = TRUE}
spatial_sf <- spatial %>% sf::st_as_sf(coords = c("long", "lat"), crs = 4326) 
station_vec <- st_geometry(spatial_sf)
time_vec <- unique(temporal$time)
var_vec <- spatial_sf$name
d <- st_dimensions(station = station_vec, time = time_vec)
d
```

However, the issue comes when casting the variables into an array of the same dimension:

```{r echo = TRUE, eval = FALSE}
temporal$var1 %>% matrix(byrow = TRUE, nrow = 2, ncol = 5) %>% array(dim = dim(d))
```

Due to the missingness on 2023 March 21st for id2, the array reuses the first value of the variable to fill in the dimension (the missingness at (2,1) is not flagged and 3 is reused to fill in the dimension).

This causes an issue when multiple stations with different recording periods present in the data, commonly found in the weather data.

The \pkg{sftime} package doesn't have this issue.

Consider the climate data used in the \pkg{cubble} package, \code{climate_aus}. It contains daily precipitation, minimum and maximum temperature for 639 weather stations across Australia in 2020. A `sftime` object can be created from the spatial and temporal aspect of the data as follows:

```{r echo = TRUE}
library(sftime)

climate_spatial <- climate_aus %>% select(id:wmo_id)
climate_temporal <- climate_aus %>% face_temporal() %>% as_tibble()

data_sftime <- left_join(climate_spatial, climate_temporal,  
                         by = "id", multiple = "all") %>% 
  st_as_sftime(coords = c("long", "lat"))
data_sftime
```

An issue here is the object size created of `data_sftime`. whichis `r `object.size(data_sftime)/object.size(climate_aus)` times the size of \code{climate_aus}

```{r echo = TRUE}
object.size(data_sftime)
object.size(climate_aus)
```

This is the initial issue raised from this paper where a conventional left join of long temporal data with spatial object can create the large spatio-temporal data by repeating the spatial geometry at every time stamp.

### 2023-03-25 5-3-river-level-rainfall.Rmd
language edit'
some outlines on comparing between cubble/ sftime/stars

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology [@bom] in relation to weather station rainfall from NOAA's climate data (\code{climate}). The goal is to match water gauges with nearby weather stations, spatially and temporally, using 2020 measurements using \code{match_sites()} function.
<!--The Bureau of Meteorology collects [water data](http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full) from river gauges. The collected variables include: electrical conductivity, turbidity, watercourse discharge, watercourse level, and water temperature. We expect that the water level should be related to the precipitation available from climate data, since rainfall will raise the water level in the river. Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria, and we will try to match water gauges with nearby weather stations, taking into account both spatial and temporal information from 2020.-->

@@

climate <- climate_full |>
  filter(between(stringr::str_sub(id, 7, 8), 76, 90)) |>
  face_temporal() |>
  filter(lubridate::year(date) == 2020) |>
  face_spatial() |>
  mutate(type = "climate")

>>>>

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology [@bom] in relation to weather station rainfall from NOAA's climate data (\code{climate}). The goal is to match water gauges with nearby weather stations, spatially and temporally, using the \code{match_sites()} function.
<!--The Bureau of Meteorology collects [water data](http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full) from river gauges. The collected variables include: electrical conductivity, turbidity, watercourse discharge, watercourse level, and water temperature. We expect that the water level should be related to the precipitation available from climate data, since rainfall will raise the water level in the river. Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria, and we will try to match water gauges with nearby weather stations, taking into account both spatial and temporal information from 2020.-->

@@

climate <- climate_full |>
  filter(between(as.numeric(stringr::str_sub(id, 7, 8)), 76, 90)) |>
  face_temporal() |>
  filter(lubridate::year(date) == 2020) |>
  face_spatial() |>
  mutate(type = "climate")

### 2023-03-25 jss/main.tex
some outlines on comparing between cubble/ sftime/stars

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cb <- as_cubble(
+   list(spatial = lga, temporal = covid),
+   key = lga, index = date, coords = c(cent_long, cent_lat)
+   )
\end{CodeInput}
\begin{CodeOutput}
! Some sites in the temporal table don't have spatial information
\end{CodeOutput}
\begin{CodeOutput}
! Some sites in the spatial table don't have temporal information
\end{CodeOutput}
\begin{CodeOutput}
! Use argument `output = "unmatch"` to check on the unmatched key
\end{CodeOutput}
\begin{CodeOutput}
Joining, by = "lga"
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pair <- as_cubble(
+   list(spatial = lga, temporal = covid),
+   key = lga, index = date, coords = c(cent_long, cent_lat),
+   output = "unmatch"
+   )
R> 
R> pair
\end{CodeInput}
\begin{CodeOutput}
$paired
# A tibble: 2 x 2
  spatial             temporal    
  <chr>               <chr>       
1 Kingston (C) (Vic.) Kingston (C)
2 Latrobe (C) (Vic.)  Latrobe (C)

@@

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- historical_tmax |>
+   filter(between(stringr::str_sub(id, 7, 8), 46, 90))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> tmax |> 
+   face_spatial() |> 
+   mutate(n = nrow(ts)) |>
+   arrange(n) |>
+   pull(n) |> 
+   head(10)
\end{CodeInput}
\begin{CodeOutput}
 [1] 12 12 12 13 19 19 20 24 24 24
\end{CodeOutput}
\begin{CodeInput}
R> tmax <- tmax |> 
+   face_spatial() |> 
+   filter(nrow(ts) == 24) |>
+   face_temporal()
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], matched_var [dbl]
  id          name        lat  long type   dist group ts       n_match
  <chr>       <chr>     <dbl> <dbl> <chr> <dbl> <int> <list>     <int>
1 405234      SEVEN CR~ -36.9  146. river  6.15     5 <tibble>      21
2 404207      HOLLAND ~ -36.6  146. river  8.54    10 <tibble>      21
3 ASN00082042 strathbo~ -36.8  146. clim~  6.15     5 <tibble>      21
4 ASN00082170 benalla ~ -36.6  146. clim~  8.54    10 <tibble>      21
5 230200      MARIBYRN~ -37.7  145. river  6.17     6 <tibble>      19
# ... with 3 more rows
\end{CodeOutput}
\end{CodeChunk}

>>>>

\hypertarget{comparison-with-existing-packages}{%
\subsection{Comparison with existing packages}\label{comparison-with-existing-packages}}

The development on spatio-temporal data structure in R has witnessed great changes from the initial \pkg{spacetime} \citep{spacetime} package. New packages, such as \pkg{stars} \citep{stars} and \pkg{sftime} \citep{sftime}, have {[}\ldots{]}.

Below compare the additional benefit of cubble on top of the two packages.

The \texttt{stars} package creates a \texttt{stars} object from an array and a dimension object. In many cases, i.e.~satellite images, this is convenient, however, there are also many spatio-temporal data are formatted in 2D data frames.

This requires converting the variables into an array format.

Consider the following spatial and temporal data.

The temporal data contains two variables are recorded at two stations (\texttt{id}) from 2023 March 21st to 2023 March 25th. Notice is that id2 only records from 2023 March 22nd and variable 2 has an \texttt{NA} value on 2023 March 23rd at id2.

\begin{CodeChunk}
\begin{CodeOutput}
# A tibble: 2 x 4
     id  long   lat name     
  <dbl> <dbl> <dbl> <chr>    
1     1   130   -38 station 1
2     2   140   -44 station 2
\end{CodeOutput}
\begin{CodeOutput}
# A tibble: 9 x 4
     id time        var1  var2
  <dbl> <date>     <dbl> <dbl>
1     1 2023-03-21     3     1
2     1 2023-03-22     6     3
3     1 2023-03-23     8     4
4     1 2023-03-24     2     2
5     1 2023-03-25     5     6
# ... with 4 more rows
\end{CodeOutput}
\end{CodeChunk}

The following code can be used to create the dimension object:

\begin{CodeChunk}
\begin{CodeInput}
R> spatial_sf <- spatial %>% sf::st_as_sf(coords = c("long", "lat"), crs = 4326) 
R> station_vec <- st_geometry(spatial_sf)
R> time_vec <- unique(temporal$time)
R> var_vec <- spatial_sf$name
R> d <- st_dimensions(station = station_vec, time = time_vec)
R> d
\end{CodeInput}
\begin{CodeOutput}
        from to     offset  delta refsys point
station    1  2         NA     NA WGS 84  TRUE
time       1  5 2023-03-21 1 days   Date FALSE
                                  values
station POINT (130 -38), POINT (140 -44)
time                                NULL
\end{CodeOutput}
\end{CodeChunk}

However, the issue comes when casting the variables into an array of the same dimension:

\begin{CodeChunk}
\begin{CodeInput}
R> temporal$var1 %>% matrix(byrow = TRUE, nrow = 2, ncol = 5) %>% array(dim = dim(d))
\end{CodeInput}
\end{CodeChunk}

Due to the missingness on 2023 March 21st for id2, the array reuses the first value of the variable to fill in the dimension (the missingness at (2,1) is not flagged and 3 is reused to fill in the dimension).

This causes an issue when multiple stations with different recording periods present in the data, commonly found in the weather data.

The \pkg{sftime} package doesn't have this issue.

Consider the climate data used in the \pkg{cubble} package, \code{climate_aus}. It contains daily precipitation, minimum and maximum temperature for 639 weather stations across Australia in 2020. A \texttt{sftime} object can be created from the spatial and temporal aspect of the data as follows:

\begin{CodeChunk}
\begin{CodeInput}
R> library(sftime)
R> 
R> climate_spatial <- climate_aus %>% select(id:wmo_id)
R> climate_temporal <- climate_aus %>% face_temporal() %>% as_tibble()
R> 
R> data_sftime <- left_join(climate_spatial, climate_temporal,  
+                          by = "id", multiple = "all") %>% 
+   st_as_sftime(coords = c("long", "lat"))
R> data_sftime
\end{CodeInput}
\begin{CodeOutput}
Spatiotemporal feature collection with 230760 features and 7 fields
Geometry type: POINT
Dimension:     XY
Bounding box:  xmin: 113.5319 ymin: -43.6578 xmax: 153.6358 ymax: -10.0511
CRS:           NA
Time column with class: 'Date'.
Ranging from 2020-01-01 to 2020-12-31.
# A tibble: 230,760 x 9
  id           elev name         wmo_id date        prcp  tmax  tmin
* <chr>       <dbl> <chr>         <dbl> <date>     <dbl> <dbl> <dbl>
1 ASN00001006   3.8 wyndham aero  95214 2020-01-01   164  38.3  25.3
2 ASN00001006   3.8 wyndham aero  95214 2020-01-02     0  40.6  30.5
3 ASN00001006   3.8 wyndham aero  95214 2020-01-03    16  39.7  27.2
4 ASN00001006   3.8 wyndham aero  95214 2020-01-04     0  38.2  27.3
5 ASN00001006   3.8 wyndham aero  95214 2020-01-05     2  39.3  26.7
# ... with 230,755 more rows, and 1 more variable: geometry <POINT>
\end{CodeOutput}
\end{CodeChunk}

An issue here is the object size created of \texttt{data\_sftime}. whichis \texttt{r}object.size(data\_sftime)/object.size(climate\_aus)` times the size of \code{climate_aus}

\begin{CodeChunk}
\begin{CodeInput}
R> object.size(data_sftime)
\end{CodeInput}
\begin{CodeOutput}
114549848 bytes
\end{CodeOutput}
\begin{CodeInput}
R> object.size(climate_aus)
\end{CodeInput}
\begin{CodeOutput}
8641608 bytes
\end{CodeOutput}
\end{CodeChunk}

This is the initial issue raised from this paper where a conventional left join of long temporal data with spatial object can create the large spatio-temporal data by repeating the spatial geometry at every time stamp.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cb <- as_cubble(
+   list(spatial = lga, temporal = covid),
+   key = lga, index = date, coords = c(cent_long, cent_lat)
+   )
\end{CodeInput}
\begin{CodeOutput}
! Some sites in the temporal table don't have spatial information
\end{CodeOutput}
\begin{CodeOutput}
! Some sites in the spatial table don't have temporal information
\end{CodeOutput}
\begin{CodeOutput}
! Use argument `output = "unmatch"` to check on the unmatched key
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pair <- as_cubble(
+   list(spatial = lga, temporal = covid),
+   key = lga, index = date, coords = c(cent_long, cent_lat),
+   output = "unmatch"
+   )
R> 
R> pair
\end{CodeInput}
\begin{CodeOutput}
$paired
# A tibble: 80 x 2
  spatial        temporal      
  <chr>          <chr>         
1 Alpine (S)     Alpine (S)    
2 Ararat (RC)    Ararat (RC)   
3 Ballarat (C)   Ballarat (C)  
4 Banyule (C)    Banyule (C)   
5 Bass Coast (S) Bass Coast (S)
# ... with 75 more rows

@@

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- historical_tmax |>
+   filter(between(as.numeric(stringr::str_sub(id, 7, 8)), 46, 90))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> tmax |> 
+   face_spatial() |> 
+   mutate(n = nrow(ts)) |>
+   arrange(n) |>
+   pull(n) |> 
+   head(10)
\end{CodeInput}
\begin{CodeOutput}
 [1] 24 24 12 24 24 24 24 24 24 24
\end{CodeOutput}
\begin{CodeInput}
R> tmax <- tmax |> 
+   face_spatial() |> 
+   filter(nrow(ts) == 24) |>
+   face_temporal()
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], matched_var [dbl]
  id          name        lat  long type   dist group ts       n_match
  <chr>       <chr>     <dbl> <dbl> <chr> <dbl> <int> <list>     <int>
1 230200      MARIBYRN~ -37.7  145. river  6.17     6 <tibble>      19
2 404207      HOLLAND ~ -36.6  146. river  8.54    10 <tibble>      21
3 405234      SEVEN CR~ -36.9  146. river  6.15     5 <tibble>      21
4 406213      CAMPASPE~ -37.0  145. river  1.84     1 <tibble>      18
5 ASN00082042 strathbo~ -36.8  146. clim~  6.15     5 <tibble>      21
# ... with 3 more rows
\end{CodeOutput}
\end{CodeChunk}
